---
subtitle: ""
format:
  pdf:
    toc: false
    number-sections: false
    geometry:
      - margin=0.7in
      - top=0.9in
    fontsize: 9pt
    header-includes:
      - \usepackage{fancyhdr}
      - \pagestyle{fancy}
      - \fancyhf{}
      - \fancyhead[L]{\small NSQIP CUSUM — Executive Summary}
      - \fancyhead[R]{\small\thepage}
      - \renewcommand{\headrulewidth}{0.4pt}
      - \usepackage{float}
      - \usepackage{longtable}
      - \usepackage{needspace}
      - \floatplacement{figure}{H}
      - \usepackage{colortbl}
      - \definecolor{sigred}{HTML}{CB181D}
      - \definecolor{okgreen}{HTML}{238B45}
      - \setlength{\parskip}{4pt}
      - \setlength{\parindent}{0pt}
params:
  specialty: "General Surgery"
  division: ""
  # A prefix, not an exact name — see resolve_case_file() in R/data_processing.R
  data_file: "data/Case_Details_Report.xlsx"
  site_sar_file: "data/SAR_Site_Summary.xlsx"
  surgeon_mapping_file: "data/surgeon_division_mapping.csv"
  benchmark_type: "site_expected"
  specialties: ["General Surgery", "Vascular", "Thoracic", "Plastics"]
  odds_ratio: 2.0
  target_arl: 1500
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

# The executive summary is a selection and layout layer over the full report.
# It computes nothing the full report does not already compute — every number
# here comes from the same build_triage() / build_flag_procedures() /
# build_flag_caselist() calls, against the same cached input bundle. If a
# number differs between the two documents, that is a bug, not a difference of
# method.

source("R/version.R")
source("R/benchmarks.R")
source("R/data_processing.R")
source("R/cusum_functions.R")
source("R/triage.R")
source("R/load_report_data.R")
source("R/output_names.R")

library(knitr)
library(kableExtra)

options(readxl.show_progress = FALSE)
```

```{r load-data}
#| include: false

bundle <- load_report_data(
  data_file            = params$data_file,
  site_sar_file        = params$site_sar_file,
  surgeon_mapping_file = params$surgeon_mapping_file,
  benchmark_type       = params$benchmark_type,
  specialties          = unlist(params$specialties)
)

case_data       <- bundle$case_data
benchmark_rates <- bundle$benchmark_rates
targeted_p0     <- targeted_for_mode(bundle, params$benchmark_type)

spec <- params$specialty
div  <- if (nchar(params$division) > 0) params$division else NULL

spec_data <- case_data |> filter(specialty == spec)
if (!is.null(div)) spec_data <- spec_data |> filter(division == div)

n_cases  <- nrow(spec_data)
date_min <- min(spec_data$op_date, na.rm = TRUE)
date_max <- max(spec_data$op_date, na.rm = TRUE)

report_name <- if (!is.null(div)) paste0(spec, " — ", div) else spec

bench_desc <- if (params$benchmark_type == "site_expected" &&
                  bundle$site_sar_available) {
  "site risk-adjusted expected rates (SAR/ISAR)"
} else {
  "national observed rates (SAR)"
}
```

```{r triage-core}
#| include: false

triage <- build_triage(
  data            = case_data,
  spec            = spec,
  div             = div,
  benchmark_rates = benchmark_rates,
  targeted_rates  = targeted_p0,
  odds_ratio      = params$odds_ratio,
  target_arl      = params$target_arl
)
triage <- annotate_carryover(triage, spec, div)
triage <- annotate_composite_overlap(triage, case_data, spec, div)

flagged <- if (!is.null(triage)) triage |> filter(tier > 0) else NULL
n_flags <- if (is.null(flagged)) 0L else nrow(flagged)

# The detection floor is what makes "nothing flagged" readable, and for most
# scopes it is the substance of the page. Built here rather than at the point
# of printing so the bottom-line text can promise it only when it exists.
floor_tbl <- if (is.null(triage)) NULL else {
  triage |>
    filter(!is.na(need_events)) |>
    arrange(desc(fold)) |>
    transmute(
      Complication     = complication,
      Exp              = expected,
      Obs              = observed,
      `Events to flag` = need_events,
      `= fold excess`  = fold
    )
}
has_floor <- !is.null(floor_tbl) && nrow(floor_tbl) > 0
```

\vspace{-2.5em}

## `r report_name` {.unnumbered}

**Chart review summary** · `r format(date_min, "%d %b %Y")` – `r format(date_max, "%d %b %Y")` · `r n_cases` cases · benchmark: `r bench_desc`

\vspace{-0.4em}
\hrulefill

```{r bottom-line}
#| results: asis

if (n_flags == 0) {
  cat("### Bottom line: nothing met the review threshold\n\n")
  cat("No complication was elevated enough, or clustered recently enough, to",
      "warrant pulling charts this period.\n\n")
  cat("**Read this as \"nothing above the detection floor\", not as \"nothing",
      "to see\".** With", n_cases, "cases in this window some complications",
      "are too rare to flag at any plausible rate",
      if (has_floor) "— the table below shows what each one would have needed."
      else "regardless of performance.",
      "\n\n")
} else {
  n_t1 <- sum(flagged$tier == 1)
  cat("### Bottom line: ", n_flags, " complication",
      if (n_flags > 1) "s" else "", " on the worklist\n\n", sep = "")
  cat("The worklist is below, with the procedures and the cases behind it.",
      if (n_t1 > 0) {
        paste0("Start with the ", n_t1, " Tier 1 item",
               if (n_t1 > 1) "s" else "",
               ": rate elevated *and* events clustered recently.")
      } else {
        "Nothing reached Tier 1 — no flag has both an elevated rate and a recent cluster."
      },
      "\n\n")
}
```

```{r worklist}
#| results: asis

if (n_flags > 0) {
  cat("### The worklist\n\n")

  tbl <- flagged |>
    transmute(
      Priority     = tier_label,
      Complication = complication,
      Obs          = observed,
      Exp          = expected,
      `O/E`        = oe,
      `p`          = p_value,
      CUSUM        = ifelse(cusum_signal, "yes", "--"),
      Status       = status
    )

  k <- kable(tbl, format = "latex", booktabs = TRUE, linesep = "",
             align = "llrrrrcl") |>
    kable_styling(latex_options = c("hold_position"), font_size = 8)
  t1 <- which(flagged$tier == 1)
  if (length(t1) > 0) k <- k |> row_spec(t1, bold = TRUE)
  print(k)
  cat("\n")

  cat("\\footnotesize A complication is listed when its cumulative rate is",
      "elevated (at least", TRIAGE_MIN_EVENTS, "events and one-sided",
      paste0("$p < ", TRIAGE_ALPHA, "$"),
      "against expected) or the CUSUM shows recent clustering.",
      "Clustering with observed at or below expected is not listed.",
      "\\normalsize\n\n")

  # Where two flags are one problem. Load-bearing: without it a division opens
  # two chart reviews on the same patients.
  notes <- flagged |> filter(!is.na(overlap_note))
  if (nrow(notes) > 0) {
    for (i in seq_len(nrow(notes))) {
      cat("- **", notes$complication[i], ":** ", notes$overlap_note[i],
          "\n", sep = "")
    }
    cat("\n")
  }
}
```

```{r where-events-sit}
#| results: asis

proc_break <- build_flag_procedures(case_data, spec, div, triage)

if (!is.null(proc_break) && nrow(proc_break) > 0) {
  # A composite adding no patients would repeat the explaining flag's table
  # verbatim, which reads as a second finding.
  redundant <- if ("overlap_full" %in% names(triage)) {
    triage$complication[which(triage$overlap_full)]
  } else character(0)
  proc_break <- proc_break |> filter(!(complication %in% redundant))
}

if (!is.null(proc_break) && nrow(proc_break) > 0) {
  cat("### Where the events sit\n\n")
  cat("Ranked by event count, not rate. The rollup row is the comparison —",
      "a procedure only stands out against the rest of the service.\n\n")

  for (comp in unique(proc_break$complication)) {
    sub <- proc_break |> filter(complication == comp)
    ptbl <- sub |>
      transmute(CPT = cpt, Procedure = description,
                Events = events, Cases = cases, `Rate %` = rate_pct)
    cat("**", comp, " — events by procedure**\n\n", sep = "")
    k <- kable(ptbl, format = "latex", booktabs = TRUE, linesep = "",
               align = "llrrr") |>
      kable_styling(latex_options = c("hold_position"), font_size = 8)
    other_row <- which(sub$is_other)
    if (length(other_row) > 0) k <- k |> row_spec(other_row, italic = TRUE)
    print(k)
    cat("\n")
  }
}
```

```{r detection-floor}
#| results: asis

if (has_floor) {
  cat("### What this report could have detected\n\n")
  cat("What each complication would have needed to flag, given", n_cases,
      "cases. A high fold value means absence of a flag says little.\n\n")
  print(
    kable(floor_tbl, format = "latex", booktabs = TRUE, linesep = "",
          align = "lrrrr") |>
    kable_styling(latex_options = c("hold_position"), font_size = 7)
  )
  cat("\n")
}
```

```{r cases-to-pull}
#| results: asis

caselist <- build_flag_caselist(case_data, spec, div, triage)

if (!is.null(caselist) && nrow(caselist) > 0) {
  cat("\\needspace{9\\baselineskip}\n\n")
  cat("### Cases to pull\n\n")
  cat("Every case carrying a flagged complication, across the full report",
      "window, so the count reconciles with the Obs column above. PATOS",
      "(present at time of surgery) events are excluded.\n\n")

  kable(caselist, format = "latex", booktabs = TRUE, linesep = "",
        align = c("l", "l", "l", "l", "c", "r", "l"),
        longtable = TRUE) |>
    kable_styling(latex_options = "hold_position", font_size = 7) |>
    print()
  cat("\n")
}
```

\vspace{0.4em}
\hrulefill

\footnotesize
**This is a summary.** CUSUM charts, prior SAR performance, the monthly
complication dashboard, procedure mix and the full complication case list are
in the accompanying report,
`r latex_escape_filename(paste0(scope_base_name(spec, params$division), ".pdf"))`.
No number here is computed differently from the full report.

Generated `r format(Sys.time(), "%Y-%m-%d %H:%M")` · `r nsqip_version_label()``r if (!is.null(div)) paste0(" · Division: ", div)`
\normalsize
