---
title: "NSQIP CUSUM Outcome Monitoring Report"
subtitle: "`r if (nchar(params$division) > 0) paste0(params$specialty, ' \\u2014 ', params$division) else params$specialty`"
author: "Surgical Quality Improvement"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: false
    geometry:
      - margin=0.75in
    fontsize: 10pt
    header-includes:
      - \usepackage{fancyhdr}
      - \pagestyle{fancy}
      - \fancyhead[L]{NSQIP CUSUM Monitoring}
      - \fancyhead[R]{\nouppercase{\leftmark}}
      - \fancyfoot[C]{\thepage}
      - \usepackage{float}
      - \usepackage{longtable}
      - \floatplacement{figure}{H}
      - \usepackage{colortbl}
      - \definecolor{sigred}{HTML}{CB181D}
      - \definecolor{okgreen}{HTML}{238B45}
params:
  specialty: "General Surgery"
  division: ""
  # A prefix, not an exact name — see resolve_case_file() in R/data_processing.R
  data_file: "data/Case_Details_Report.xlsx"
  site_sar_file: "data/SAR_Site_Summary.xlsx"
  surgeon_mapping_file: "data/surgeon_division_mapping.csv"
  benchmark_type: "site_expected"
  specialties: ["General Surgery", "Vascular", "Thoracic", "Plastics"]
  odds_ratio: 2.0
  target_arl: 500
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

source("R/version.R")
source("R/benchmarks.R")
source("R/data_processing.R")
source("R/cusum_functions.R")
source("R/triage.R")
source("R/load_report_data.R")

library(knitr)
library(kableExtra)

options(readxl.show_progress = FALSE)
```

```{r load-data}
#| include: false

# Shared with the slide renderer; see R/load_report_data.R
bundle <- load_report_data(
  data_file            = params$data_file,
  site_sar_file        = params$site_sar_file,
  surgeon_mapping_file = params$surgeon_mapping_file,
  benchmark_type       = params$benchmark_type,
  specialties          = unlist(params$specialties)
)

case_data          <- bundle$case_data
benchmark_rates    <- bundle$benchmark_rates
targeted_data      <- bundle$targeted_data
ot_data            <- bundle$ot_data
site_sar_available <- bundle$site_sar_available

# Risk-adjusted targeted rates belong only in site_expected mode
targeted_p0 <- targeted_for_mode(bundle, params$benchmark_type)

spec <- params$specialty
div  <- if (nchar(params$division) > 0) params$division else NULL

# Filter to this specialty (and division if specified)
spec_data <- case_data |> filter(specialty == spec)
if (!is.null(div)) {
  spec_data <- spec_data |> filter(division == div)
}

n_cases  <- nrow(spec_data)
date_min <- min(spec_data$op_date, na.rm = TRUE)
date_max <- max(spec_data$op_date, na.rm = TRUE)

# Display name for titles
report_name <- if (!is.null(div)) paste0(spec, " — ", div) else spec

# Benchmark description for text
bench_desc <- if (params$benchmark_type == "site_expected" && site_sar_available) {
  "site-specific risk-adjusted expected rates from the most recent SAR/ISAR"
} else {
  "national observed rates from the most recent SAR"
}
```

\newpage

# Overview

This report presents Bernoulli CUSUM charts for monitoring surgical outcomes
in **`r report_name`** using site-level NSQIP case data.

**Report period:** `r format(date_min, "%B %d, %Y")` to `r format(date_max, "%B %d, %Y")`
\
**Total cases:** `r n_cases`
`r if (!is.null(div)) paste0("\\\\\n**Division:** ", div, " (within ", spec, ")")` 
\
**Benchmark:** `r bench_desc`

## Methodology

Each chart tracks one complication using a Bernoulli CUSUM:

- **p~0~ (benchmark rate):** `r if (params$benchmark_type == "site_expected") "Risk-adjusted expected rate for this site's case mix, derived from the SAR hierarchical model. Where unavailable, the national observed rate is used." else "National observed rate for this specialty from the SAR."``r if (!is.null(div)) " Benchmarks are assigned per case: the matching targeted procedure model (e.g. Colectomy, Proctectomy) where one exists, otherwise the specialty model."`
- **p~1~ (detection target):** Rate corresponding to an odds ratio of
  `r params$odds_ratio`$\times$ relative to p~0~.
- **h (decision boundary):** Simulated at render time for ARL~0~ $\approx$
  `r params$target_arl` cases against this chart's own p~0~ mix. Each chart
  subtitle reports the ARL achieved and how many events in quick succession
  trip it.
- **Signal ($\blacktriangle$):** CUSUM crossed h; chart resets to zero.
- **Events (|):** Orange tick marks show individual complication occurrences.
- **Title color:** \textcolor{sigred}{Red} = SAR "Needs Improvement";
  \textcolor{okgreen}{Green} = SAR "Exemplary"; Black = "As Expected" or no SAR model.

**Key limitations:** `r if (params$benchmark_type == "site_expected") "While the benchmark rates are risk-adjusted, the CUSUM chart itself monitors the raw (unadjusted) event stream against those expected rates. Changes in case mix during the monitoring period are not dynamically accounted for." else "These charts use unadjusted rates. Case mix differences from the national average will affect interpretation."``r if (!is.null(div)) " Division reports benchmark each case against the most specific SAR model available; procedures with no targeted model fall back to the specialty rate."` PATOS exclusions are applied per SAR methodology. Intended for internal QI monitoring, not formal profiling.

\newpage

# Chart Review Priorities

```{r triage}
#| results: asis

triage <- build_triage(
  data            = case_data,
  spec            = spec,
  div             = div,
  benchmark_rates = benchmark_rates,
  targeted_rates  = targeted_p0,
  odds_ratio      = params$odds_ratio,
  target_arl      = params$target_arl
)
triage <- annotate_carryover(triage, spec, div)
triage <- annotate_composite_overlap(triage, case_data, spec, div)

cat("A complication is listed for review only when the cumulative rate is",
    "elevated (at least", TRIAGE_MIN_EVENTS, "events and one-sided p <",
    TRIAGE_ALPHA, "against expected) or the CUSUM shows recent clustering.",
    "A CUSUM signal on its own, where observed events are at or below",
    "expected, reflects clustering rather than excess and is not listed.\n\n")

flagged <- if (!is.null(triage)) triage |> filter(tier > 0) else NULL

if (is.null(flagged) || nrow(flagged) == 0) {
  cat("**No complication met the review threshold this period.**\n\n")
} else {
  tbl <- flagged |>
    transmute(
      Priority       = tier_label,
      Complication   = complication,
      Obs            = observed,
      Exp            = expected,
      `O/E`          = oe,
      `p`            = p_value,
      `CUSUM`        = ifelse(cusum_signal, "yes", "--"),
      Status         = status
    )
  print(kable(tbl, format = "latex", booktabs = TRUE, linesep = "",
              caption = "Complications meeting the review threshold") |>
        kable_styling(latex_options = c("hold_position"), font_size = 9))
  cat("\n\n")

  # Composite overlap: say when two flags are one problem, so a division does
  # not open two chart reviews on the same patients.
  notes <- flagged |> filter(!is.na(overlap_note))
  if (nrow(notes) > 0) {
    for (i in seq_len(nrow(notes))) {
      cat("**", notes$complication[i], ":** ", notes$overlap_note[i],
          "\n\n", sep = "")
    }
  }
}

# Where the events sit. A flag names a complication; this names the procedures
# to pull charts on.
proc_break <- build_flag_procedures(case_data, spec, div, triage)

if (!is.null(proc_break) && nrow(proc_break) > 0) {

  # A composite that adds no patients would repeat the explaining flag's table
  # verbatim, which reads as a second finding.
  redundant <- if ("overlap_full" %in% names(triage)) {
    triage$complication[which(triage$overlap_full)]
  } else character(0)
  proc_break <- proc_break |> filter(!(complication %in% redundant))
}

if (!is.null(proc_break) && nrow(proc_break) > 0) {
  cat("## Where the events sit\n\n")
  cat("For each flagged complication, the CPTs carrying the most events.",
      "Rows are ordered by event count, not rate, and denominators are shown",
      "as they are: at these volumes a high rate on a handful of cases is",
      "common and usually means nothing. The rollup row is the comparison —",
      "a procedure only stands out relative to the rest of the service.\n\n")

  for (comp in unique(proc_break$complication)) {
    sub <- proc_break |> filter(complication == comp)
    ptbl <- sub |>
      transmute(
        CPT           = cpt,
        Procedure     = description,
        Events        = events,
        Cases         = cases,
        `Rate %`      = rate_pct
      )
    k <- kable(ptbl, format = "latex", booktabs = TRUE, linesep = "",
               caption = paste0(comp, " — events by procedure")) |>
      kable_styling(latex_options = c("hold_position"), font_size = 8)
    other_row <- which(sub$is_other)
    if (length(other_row) > 0) k <- k |> row_spec(other_row, italic = TRUE)
    print(k)
    cat("\n\n")
  }
}

# Sensitivity floor, so "nothing flagged" is interpretable
if (!is.null(triage)) {
  floor_tbl <- triage |>
    filter(!is.na(need_events)) |>
    arrange(desc(fold)) |>
    transmute(
      Complication = complication,
      Exp          = expected,
      Obs          = observed,
      `Events to flag` = need_events,
      `= fold excess`  = fold
    )
  cat("## What this report could have detected\n\n")
  cat("With", n_cases,
      "cases in this window, the counts below are what each",
      "complication would have needed to flag. A complication with a high",
      "fold value is one where absence of a flag says little.\n\n")
  print(kable(floor_tbl, format = "latex", booktabs = TRUE, linesep = "",
              caption = "Detection floor for this reporting window") |>
        kable_styling(latex_options = c("hold_position"), font_size = 8))
  cat("\n\n")
}
```

\newpage

# SAR Context — Prior Performance

```{r sar-context-table}
#| results: asis

if (site_sar_available) {
  
  site_sar_data <- parse_site_sar(params$site_sar_file) |>
    filter(specialty == spec)
  
  sar_model_label <- if (!is.null(div)) {
    paste0(spec, " SAR Model (specialty-level, not division-specific)")
  } else {
    "Most Recent SAR Results"
  }
  
  if (nrow(site_sar_data) > 0) {
    sar_tbl <- site_sar_data |>
      mutate(
        obs_pct = round(observed_rate_sar * 100, 2),
        exp_pct = round(expected_rate * 100, 2),
        oe      = round(odds_ratio_sar, 2),
        ci      = paste0("(", round(ci_lower, 2), "–", round(ci_upper, 2), ")"),
        pctl    = round(adj_percentile)
      ) |>
      select(
        Complication  = complication,
        `N (SAR)`     = n_cases_sar,
        `Obs %`       = obs_pct,
        `Exp %`       = exp_pct,
        `O/E`         = oe,
        `95% CI`      = ci,
        Pctl          = pctl,
        Assessment    = assessment
      )
    
    kable(sar_tbl, format = "latex", booktabs = TRUE, align = "lrcccclc",
          caption = paste0(report_name, " — ", sar_model_label)) |>
      kable_styling(latex_options = c("hold_position", "scale_down"),
                    font_size = 8) |>
      column_spec(8, bold = TRUE) |>
      print()
    
    cat("\n\n")
    
    # Count needs improvement
    n_ni <- sum(site_sar_data$assessment == "Needs Improvement", na.rm = TRUE)
    n_ex <- sum(site_sar_data$assessment == "Exemplary", na.rm = TRUE)
    
    if (n_ni > 0) {
      ni_comps <- site_sar_data |>
        filter(assessment == "Needs Improvement") |>
        pull(complication)
      cat(paste0("**Needs Improvement (", n_ni, "):** ",
                 paste(ni_comps, collapse = ", "), "\n\n"))
    }
    if (n_ex > 0) {
      ex_comps <- site_sar_data |>
        filter(assessment == "Exemplary") |>
        pull(complication)
      cat(paste0("**Exemplary (", n_ex, "):** ",
                 paste(ex_comps, collapse = ", "), "\n\n"))
    }
    
  } else {
    cat("No site SAR data available for this specialty.\n\n")
  }
} else {
  cat("Site SAR file not provided. Using national benchmarks only.\n\n")
}
```

\newpage

# Current Observed Rates vs. Benchmarks

```{r summary-table}
obs_summary <- observed_rates_summary(case_data, spec, div)

bench_for_spec <- benchmark_rates |>
  filter(specialty == spec) |>
  select(complication, p0_pct, benchmark_source, national_rate_pct)

summary_tbl <- obs_summary |>
  left_join(bench_for_spec, by = "complication") |>
  mutate(
    benchmark_pct = round(p0_pct, 2),
    ratio = ifelse(p0_pct > 0, round(observed_rate_pct / p0_pct, 2), NA),
    bench_type = ifelse(grepl("Site", benchmark_source), "Site Exp", "National")
  ) |>
  select(
    Complication   = complication,
    `N`            = n_cases,
    Events         = n_events,
    `Obs %`        = observed_rate_pct,
    `Benchmark %`  = benchmark_pct,
    `O/B`          = ratio,
    Type           = bench_type
  )

kable(summary_tbl, format = "latex", booktabs = TRUE, align = "lrrcccl",
      caption = paste0("Current Observed vs. Benchmark — ", report_name)) |>
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 9) |>
  footnote(
    general = "O/B = Observed/Benchmark ratio. Type: Site Exp = risk-adjusted expected rate; National = national observed rate.",
    general_title = "", footnote_as_chunk = TRUE
  )
```

\newpage

# Monthly Complication Dashboard

```{r dashboard}
#| results: asis

dashboard <- build_dashboard(case_data, spec, div, benchmark_rates)

if (!is.null(dashboard)) {
  
  dtbl <- dashboard$table
  
  cat(paste0(
    "Monthly event counts for ", report_name, " (",
    dashboard$months[1], "–", dashboard$months[length(dashboard$months)],
    " ", format(date_min, "%Y"),
    "). PATOS events excluded. SAR \\% shows site-level risk-adjusted ",
    "expected rates where available.\n\n"
  ))
  
  # Format display table
  dtbl_display <- dtbl
  month_cols <- dashboard$months
  
  # Convert month/total columns to character so we control display
  for (mc in c(month_cols, "Total")) {
    dtbl_display[[mc]] <- as.character(dtbl_display[[mc]])
  }
  
  # Format Rate % and SAR %: blank if NA, otherwise one decimal
  dtbl_display$`Rate %` <- ifelse(
    is.na(dtbl$`Rate %`), "",
    format(dtbl$`Rate %`, nsmall = 1)
  )
  dtbl_display$`SAR %` <- ifelse(
    is.na(dtbl$`SAR %`), "",
    format(dtbl$`SAR %`, nsmall = 1)
  )
  
  k <- kable(dtbl_display, format = "latex", booktabs = TRUE,
        align = c("l", rep("c", length(month_cols)), "r", "r", "r"),
        caption = paste0("Monthly Complication Dashboard — ", report_name),
        linesep = "") |>
    kable_styling(latex_options = c("hold_position", "scale_down"),
                  font_size = 7) |>
    row_spec(1, bold = TRUE, background = "#E8E8E8") |>
    row_spec(2, bold = TRUE)
  
  # Add group row packing
  for (g in names(dashboard$pack_rows)) {
    start_row <- dashboard$pack_rows[[g]][1]
    end_row   <- dashboard$pack_rows[[g]][2]
    n_rows    <- end_row - start_row + 1
    k <- k |> pack_rows(g, start_row, end_row,
                         bold = TRUE, italic = FALSE,
                         hline_before = TRUE, hline_after = FALSE)
  }
  
  print(k)
  
}
```

\newpage

# CUSUM Charts

```{r generate-charts}
#| results: hide

charts <- generate_specialty_charts(
  data       = case_data,
  spec       = spec,
  rates      = benchmark_rates,
  odds_ratio = params$odds_ratio,
  div        = div,
  targeted_rates = targeted_p0,
  target_arl = params$target_arl
)
```

```{r display-charts}
#| fig.width: 7.5
#| fig.height: 3.5
#| out.width: "100%"
#| results: asis

if (length(charts) == 0) {
  cat("No charts generated for this specialty.\n")
} else {
  for (i in seq_along(charts)) {
    print(charts[[i]])
    cat("\n\n")
  }
}
```

\newpage

# O/E Trends Across SAR Periods

```{r oe-trends}
#| fig.width: 7.5
#| fig.height: 2.5
#| out.width: "100%"
#| results: asis

if (!is.null(ot_data)) {
  spec_ot <- ot_data |> filter(specialty == spec)
  
  if (nrow(spec_ot) > 0) {
    
    # Describe what data is being shown
    if (!is.null(div)) {
      cat(paste0("Historical O/E ratios from the **", spec,
                 "** SAR model (specialty-level; division-specific data not available for ",
                 div, ").\n\n"))
    } else {
      cat("Historical O/E ratios across SAR periods for complications with\n")
      cat("available trend data.\n\n")
    }
    
    cat("Red dots indicate periods where the site was flagged as a high outlier.\n\n")
    
    # Show trends for complications that had signals or are Needs Improvement
    comps_of_interest <- unique(spec_ot$complication)
    
    # Prioritize: show Needs Improvement and any with high outlier history
    ni_comps <- benchmark_rates |>
      filter(specialty == spec, assessment == "Needs Improvement") |>
      pull(complication)
    
    high_outlier_comps <- spec_ot |>
      filter(is_outlier_high) |>
      pull(complication) |>
      unique()
    
    priority_comps <- unique(c(ni_comps, high_outlier_comps))
    
    if (length(priority_comps) > 0) {
      for (comp in priority_comps) {
        comp_ot <- spec_ot |> filter(complication == comp) |> arrange(period)
        if (nrow(comp_ot) >= 2) {
          trend_plot <- plot_oe_trend(comp_ot, comp)
          if (!is.null(trend_plot)) {
            print(trend_plot)
            cat("\n\n")
          }
        }
      }
    } else {
      cat("No complications flagged as Needs Improvement or high outlier in trend data.\n\n")
    }
  } else {
    cat("No over-time data available for this specialty.\n\n")
  }
} else {
  cat("Site SAR not provided; over-time trends unavailable.\n\n")
}
```

\newpage

# Appendix: Case Mix Summary

```{r case-mix}
asa_summary <- spec_data |>
  count(asa_class, name = "n") |>
  mutate(pct = round(n / sum(n) * 100, 1)) |>
  arrange(asa_class)

kable(asa_summary, format = "latex", booktabs = TRUE,
      col.names = c("ASA Classification", "N", "%"),
      caption = paste0("ASA Distribution — ", report_name)) |>
  kable_styling(latex_options = "hold_position", font_size = 9)
```

```{r monthly-volume}
#| fig.width: 7
#| fig.height: 2.5
#| out.width: "100%"

monthly <- spec_data |>
  mutate(month = floor_date(op_date, "month")) |>
  count(month, name = "cases")

ggplot(monthly, aes(x = month, y = cases)) +
  geom_col(fill = "#4292c6", width = 25) +
  geom_text(aes(label = cases), vjust = -0.3, size = 2.5) +
  scale_x_date(date_labels = "%b\n%Y", date_breaks = "1 month") +
  labs(title = paste0("Monthly Case Volume — ", report_name),
       x = NULL, y = "Cases") +
  theme_minimal(base_size = 9) +
  theme(panel.grid.minor = element_blank())
```

---

\newpage

# Appendix: Procedure Mix Profile

```{r procedure-mix}
#| results: asis

proc_mix <- build_procedure_mix(case_data, spec, div)

if (!is.null(proc_mix) && nrow(proc_mix) > 0) {
  
  cat(paste0(
    "Procedure categories based on NSQIP targeted module flags and CPT codes. ",
    "Complication rate reflects any individual occurrence (PATOS excluded); ",
    "morbidity composite is not counted separately.\n\n"
  ))
  
  kable(proc_mix, format = "latex", booktabs = TRUE,
        align = c("l","r","r","r","r","r"),
        caption = paste0("Procedure Mix — ", report_name)) |>
    kable_styling(latex_options = "hold_position", font_size = 8) |>
    print()
  
} else {
  cat("Procedure mix data not available.\n\n")
}
```

```{r targeted-sar}
#| results: asis

if (!is.null(targeted_data)) {
  
  tgt_summary <- build_targeted_summary(targeted_data, case_data, spec, div)
  div_proc_rates <- build_division_procedure_rates(
    case_data, spec, div, targeted_data
  )
  
  if ((!is.null(tgt_summary) && nrow(tgt_summary) > 0) ||
      !is.null(div_proc_rates)) {
    
    cat("\\newpage\n\n")
    cat("# Appendix: Targeted Procedure Benchmarks (SAR)\n\n")
    
    scope_note <- if (!is.null(div)) {
      paste0(
        "Procedure-specific outcomes for the **", div,
        "** division compared against site-level targeted SAR benchmarks. ",
        "SAR benchmarks reflect all cases across the entire site (not just this division) ",
        "and are risk-adjusted. Division observed rates are unadjusted."
      )
    } else {
      paste0(
        "Site-level targeted SAR benchmarks for procedure types with dedicated ",
        "NSQIP models. Division observed rates shown alongside for comparison."
      )
    }
    cat(scope_note, "\n\n")
    
    # Render one section per targeted procedure
    for (proc in unique(tgt_summary$Procedure)) {
      
      # Site-level SAR benchmarks for this procedure
      proc_sar <- tgt_summary |> filter(Procedure == proc) |> select(-Procedure)
      
      # Division-level observed rates (if available)
      proc_div <- div_proc_rates[[proc]]
      
      if (!is.null(proc_div)) {
        # Merge: division rates alongside SAR benchmarks
        merged <- proc_sar |>
          left_join(
            proc_div |> select(Complication, `Div N` = N,
                               `Div Events` = Events, `Div Obs %` = `Obs %`),
            by = "Complication"
          )
        
        kable(merged, format = "latex", booktabs = TRUE,
              align = c("l","r","r","r","r","r","l","r","r","r"),
              caption = paste0("Targeted: ", proc,
                               " — ", report_name, " vs. Site SAR")) |>
          kable_styling(latex_options = c("hold_position", "scale_down"),
                        font_size = 7) |>
          add_header_above(c(" " = 1,
                             "Site-Level SAR (Risk-Adjusted)" = 6,
                             "Division (Unadjusted)" = 3)) |>
          print()
      } else {
        # No division cases — show site SAR only
        kable(proc_sar, format = "latex", booktabs = TRUE,
              align = c("l","r","r","r","r","r","l"),
              caption = paste0("Targeted: ", proc,
                               " — Site-Level SAR Results")) |>
          kable_styling(latex_options = c("hold_position", "scale_down"),
                        font_size = 8) |>
          print()
      }
      
      cat("\n\n")
    }
    
  }
}
```

---

\newpage

# Appendix: Recent Cases with Complications

```{r complication-caselist}
#| results: asis

comp_caselist <- build_complication_caselist(case_data, spec, div, months = 3)

if (!is.null(comp_caselist) && nrow(comp_caselist) > 0) {
  
  # Date range for the case list
  cutoff <- max(spec_data$op_date, na.rm = TRUE) %m-% months(3)
  n_period_cases <- sum(spec_data$op_date > cutoff)
  
  cat(paste0(
    "Cases with at least one complication in the most recent 3 months ",
    "(operations after ", format(cutoff, "%B %d, %Y"), "). ",
    "**", nrow(comp_caselist), "** of **", n_period_cases,
    "** cases (", round(nrow(comp_caselist) / n_period_cases * 100, 1),
    "%) had one or more occurrences. ",
    "PATOS (present at time of surgery) events are excluded.\n\n"
  ))
  
  kable(comp_caselist, format = "latex", booktabs = TRUE,
        align = c("l","l","l","l","c","c","l"),
        caption = paste0("Cases with Complications — ", report_name,
                         " (Last 3 Months)"),
        longtable = TRUE) |>
    kable_styling(
      latex_options = c("hold_position", "repeat_header"),
      font_size = 7
    ) |>
    print()
  
} else {
  cat("No complications recorded in the most recent 3 months.\n\n")
}
```

---

*Report generated `r format(Sys.time(), "%Y-%m-%d %H:%M")` | `r nsqip_version_label()`*
\
*Benchmark: `r bench_desc``r if (!is.null(div)) paste0(" | Division: ", div)`*
